Predicting instabilities of a tuneable ring laser with an iterative map model

### 2020-11-29 Papers/Journal/Tuneable_Lasers.tex
Moved some figures
Read through and edited draft

\affil[1]{Mathematical Institute, University of Oxford, Radcliffe Observatory Quarter, Woodstock Rd, Oxford OX2 6GG, UK}
\affil[2]{Facutlty of Science, University of Ontario Institute of Technology, 2000 Simcoe St N, Oshawa, ON L1G 0C5, Canada}

@@

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions that lead to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component, we obtain an algebraic map for each component in the laser cavity, which we then iterate through to give the effect of one complete round trip. By neglecting the nonlinearity initially, we find and analyze the analytic fixed point solution of a linearly chirped Gaussian. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity, as an initial pulse evolves during many round trips of the cavity. In this nonlinear case, we find the chirp saturates, and envelope of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have many applications owing to their ultrashort pulses, a short as 50 femtoseconds \cite{chung2017}, and an ability to tune their output frequency rapidly~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes this kind of laser particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}. As well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: optical coupler, chirped fibre Bragg grating (CFBG), modulator (or saturable absorber), Er-doped fibre, and pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. Due to the high power and ultrashort duration of the pulses, the Kerr nonlinearity plays a crucial role in the dynamics within the cavity.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock is developed~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The shifts in frequencies cause the pulse to become more rectangular in the frequency domain with a linear chirp over most of the pulse. In optics, wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the Kerr nonlinearity causing the pulse to interfere with itself. New frequencies are then generated which induces the rectangular profile~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a degrade into short pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Typical cavity of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The pump laser travels clockwise around the left-hand loop (red arrows) energizing the Er-doped fibre. The main laser travels clockwise around the right-hand loop (blue arrows) passing through each component, with most of the pulse exiting the cavity through the optical coupler.}
	\label{fig:cavity}
\end{figure}

The canonical equation in nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here, $A = A(T, z)$ is the complex pulse envelope, $T$ is co-moving time, $z$ is the propagation coordinate, and $\beta_2$ and $\gamma$ are the group velocity dispersion and Kerr nonlinearity, respectively. Typically, \eqref{eq:smallnlse} is generalized to incorporate terms more specific to laser physics. A gain term and loss term are added (and sometimes higher order dispersion), to give the generalized nonlinear Schr\"odinger equation (GNLSE)~\cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain fibre, and $\alpha$ is the loss due to scattering and absorption.

The GNLSE, \eqref{eq:nlse}, has many applications in nonlinear optics and fibre optic communications, however, it is regularly generalized further to include a modulation term in the context of lasers. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. The GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity. It is assumed in \eqref{eq:nlse} that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its derivatives, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during one round trip. Often, this is the case due to the specialized, discrete components.

In this paper, we will build an iterative map model for the evolution of a laser pulse as it travels though the cavity. In Section~\ref{sec:model}, we will briefly review previous iterative mathematical models, then, we will derive our own iterative map model from \eqref{eq:nlse}. We will then non-dimensionalize the model and introduce four dimensionless parameters that govern the dynamics of the system. We will then examine the results of our model in Section~\ref{sec:results}---first solving the model analytically, by assuming the absence of the nonlinearity, then numerically solving the full nonlinear model. We will identify a sharp boundary in the parameter space separating the region of stability, and the region where modulation instability degrades the pulse. Finally, the concluding thoughts and possible ideas for future work are given in Section~\ref{sec:conclusion}.

\section{Iterative Map Model}
\label{sec:model}
To resolve some of the issues with an `average' model, it is common to use separate GNLSEs for each component and region in the laser, iteratively passing the pulse from one PDE to the next \cite{lapre2019, meng2020, oktem2010, woodward2018}. This method allows us to better encapsulate specific dynamics introduced by each component. The main drawback of this method is its extreme computational expense, since a nonlinear PDE must be numerically solved for every component and each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing that each effect is localized to its corresponding component in the cavity: essentially all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only amplified within the Er-doped fibre, \emph{et cetera}. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. We get the benefit of the fine-grain simulation over an `average' model, allowing us to reflect the specific geometry of the cavity, without much additional computational cost.

Iterative map methods have been around since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. This method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, they are phenomenologically chosen, and not necessarily obtained from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the imperative interplay of dispersion, modulation, and nonlinearity---disallowing their models to exhibit modulation instability.

We continue by deriving our own iterative map model by solving a reduced GNLSE for each component---except modulation, which we consider to be applied externally by an electro-optic modulator. Within the gain fibre, we assume all other effects are negligible, allowing us to simplify \eqref{eq:nlse} to
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A.
	\label{eq:gainexample}
\end{align}
To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}, which depends on the power of the pump laser. Thus, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right) \right)^{1/2} A
	\label{eq:gainform}
\end{align}
for the amplification of the pulse within the Er-doped fibre.

Repeating this same procedure for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \exp( - \alpha L_T )A, \\
	D(A) &= \FTi{\exp( i \omega^2 L_D\beta_2/2 ) \FT{A}}, \\
	F(A) &= A \exp( i \gamma |A|^2 L_f ),
\end{align}
where $R$ is the reflectivity of the optical coupler (depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \exp( - \alpha L_T )A$), $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the optical coupler, and $\mathcal{F}$ denotes the Fourier transform. Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. This gives the freedom of choosing the profile of the modulation how ever the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown that a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, it is common to use a saturable absorber in place of a modulator. In this case, \eqref{eq:modform} can freely be replaced by~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}},
\end{align}
where $q_0$ is the modulation depth, $|A|^2$ is the instantaneous power of the pulse, and $P_\text{sat}$ is the saturation power.

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values of the system based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem, leaving only dimensionless equations with dimensionless parameters. The magnitude of the dimensionless parameters characterizes the dominant dynamics of the system---much like how fluid flow can be described with only the Reynolds number. The structure of each process in the laser can be better understood by re-scaling the time, energy, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

Returning to the process maps \eqref{eq:gainform}--\eqref{eq:modform}, we find the non-dimensional mappings to be (after dropping the tildes)
\begin{align}
	G(A) &= \left(E_g^{-1} W \left( a E_g \textrm{e}^{E_g}\right) \right)^{1/2} A, & F(A) &= A \exp( i b |A|^2), \nonumber \\
	D(A) &= \FTi{\exp( i s^2 \omega^2 ) \FT{A}}, & L(A) &= h A, \label{eq:effects} \\
	M(A) &= \exp( -T^2 / 2 ) A, \nonumber
\end{align}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \exp( g_0 L_g ) \sim 8 \times 10^3, & 
		h &= (1 - R) \exp( - \alpha L_T) \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1, & s &= \left( \frac{\beta_2 L_D}{2 T_M^2} \right)^{1/2} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
as defined by the values in Table~\ref{tab:values}. Here, $a$ is a measure of the strength of the gain fibre, $h$ is the loss due to absorption and the optical coupler, $b$ is the ratio of the frequency shift due to the Kerr nonlinearity to the width of the pulse, and $s$ is the ratio of the effectiveness of the CFBG to the width of the pulse. Furthermore, notice that in this non-dimensional form, the modulation has no associated parameter, and each of the other processes in \eqref{eq:ndparam} have their own independent, non-dimensional parameter.

% Table of parameter values
\begin{table*}[tbp]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline\noalign{\smallskip}
		Symbol & Parameter & Value & Units \\
		\hline\noalign{\smallskip}
		$T_M$ & Characteristic modulation time & $15$--$150$ & ps \\
		$\beta_2^g L_D$ & Dispersion provided by CFBG & $10$--$2000$ & ps$^2$ \\
		$\gamma$ & Kerr nonlinearity coefficient & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$L_f$ & Length of fibre between gain and optical coupler & $0.15$--$1$ & m \\
		$L_g$ & Length of the Er-doped gain fibre & $2$--$3$ & m \\
		$\alpha$ & Loss of fibre due to scattering and absorption & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy of Er-doped gain fibre & $10^3$--$10^4$ & pJ \\
		$\beta_2^f$ & Second order fibre dispersion & $-50$--$50$ & ps$^2$/km \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		$L_T$ & Total length of fibre in the cavity & $10$--$100$ & m \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

Finally, we briefly investigate the effect the order of the components has on the dynamics. In Section~\ref{sec:model}\ref{sec:effects} we justified the order of the gain, nonlinearity, and loss, but, just assumed dispersion preceded modulation. Here, we consider the effect of swapping the dispersion and modulation components. The energy landscape (Fig.~\ref{fig:energyswitch}) is as we would expect. In the previous case of having modulation following dispersion, more energy is removed from the pulse tails. Whereas, having modulation first, more of this energy remains in the cavity. This has the effect of making the laser more powerful than in the previous case, which in turn increases the size of the unstable region. Comparing with Fig.~\ref{fig:energy}, the unstable upper-left region spreads farther rightward, and the energy is larger in the stable region---especially farther away from the origin. Lastly, once again, the energy at the origin is somewhat larger than 2.5, as predicted by \eqref{eq:maxenergy}.

@@

Portions of this work were presented at The V AMMCS International Conference in 2019, ``A new method of modelling tuneable lasers with functional composition'' \cite{metherallammcs}. BM acknowledges the support provided by the EPSRC Centre for Doctoral Training in Industrially Focused Mathematical Modelling (EP/L015803/1).

%The authors would like to thank O. Aytür for critical reading of the manuscript

>>>>

\affil[1]{Mathematical Institute, University of Oxford, Radcliffe Observatory Quarter, Woodstock Rd, Oxford OX2 6GG, UK}
\affil[2]{Faculty of Science, University of Ontario Institute of Technology, 2000 Simcoe St N, Oshawa, ON L1G 0C5, Canada}

@@

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions leading to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component in the laser cavity, we obtain an algebraic map for each component. Iterating these maps gives the total effect of one complete round trip. By neglecting the nonlinearity initially, we find a linearly chirped Gaussian to be the analytic fixed point solution, which we analyze asymptotically. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity, as the pulse evolves over many round trips of the cavity. In this nonlinear case, we find the chirp saturates, and the envelope of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have a wide range of applications, owing to their ultrashort pulses, as short as 50 femtoseconds \cite{chung2017}, and the ability to tune their output frequency rapidly over a range of over 100 nm~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes this kind of laser particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}; as well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: optical coupler, chirped fibre Bragg grating (CFBG), modulator (or saturable absorber), Er-doped fibre, and pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. Due to the high power and ultrashort duration of the pulses, the Kerr nonlinearity plays a crucial role in the dynamics within the cavity.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be redshifted, and the trailing edge to be blueshifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking manifests from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into much shorter, less intense pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Common cavity of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The pump laser travels clockwise around the left-hand loop (red arrows) energizing the Er-doped fibre. The main laser travels clockwise around the right-hand loop (blue arrows) passing through each component, with most of the pulse exiting the cavity through the optical coupler.}
	\label{fig:cavity}
\end{figure}

The canonical equation in nonlinear optics is the nonlinear Schr\"odinger equation,
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here, $A = A(T, z)$ is the complex pulse envelope, $T$ is co-moving time, $z$ is the propagation coordinate, and $\beta_2$ and $\gamma$ are the group velocity dispersion and Kerr nonlinearity, respectively. Typically, \eqref{eq:smallnlse} is generalized to incorporate terms more specific to laser physics. A gain term and loss term are added (and sometimes higher order dispersion terms), to give the generalized nonlinear Schr\"odinger equation (GNLSE)~\cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain fibre, and $\alpha$ is the loss due to scattering and absorption.

The GNLSE, \eqref{eq:nlse}, has applications in nonlinear optics and fibre optic communications, however, it is regularly generalized even further to include a modulation term. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known to the master equation. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. While, the GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity such as Fig. \ref{fig:cavity}. It is assumed in \eqref{eq:nlse} that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its generalizations, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during any one round trip. Often, this is the case due to the specialized, discrete components.

In this paper, we will build an iterative map model for the evolution of a laser pulse as it travels though the cavity. In Section~\ref{sec:model}, we will briefly review previous iterative map models, then, we will derive our own iterative map model from \eqref{eq:nlse}. We will then non-dimensionalize the model and introduce four dimensionless parameters that govern the dynamics of the system. We will then examine the results of our model in Section~\ref{sec:results}---first solving the model analytically by assuming the absence of the nonlinearity, then numerically solving the full nonlinear model. We will identify a sharp boundary in the parameter space separating the region of stability, and the region where modulation instability degrades the pulse. Finally, the concluding thoughts and ideas for possible future work are given in Section~\ref{sec:conclusion}.

\section{Iterative Map Model}
\label{sec:model}
To resolve some of the issues with an `average' model, it is common to use separate GNLSEs for each component and region in the laser, iteratively passing the pulse from one GNLSE to the next \cite{lapre2019, meng2020, oktem2010, woodward2018}. This method allows us to better encapsulate specific dynamics introduced by each component. The main drawback of this method is its extreme computational expense, since a nonlinear GNLSE must be numerically solved for every segment for each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing that each effect is localized to its corresponding component in the cavity: essentially all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only amplified within the Er-doped fibre, \emph{et cetera}. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. This reduces the computational cost, while maintaining the benefit of reflecting the specific geometry of a component-wise simulation.

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. This method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the imperative interplay of dispersion, modulation, and nonlinearity---hindering their models to exhibit modulation instability.

We continue by deriving our own iterative map model by solving a reduced GNLSE for each component---except modulation, which we consider to be applied externally by an electro-optic modulator. Within the gain fibre we assume all other effects are negligible. Any dispersion, for example, in the gain fibre pales in comparison to the dispersion introduced by the CFBG. Thus, we neglect the dispersion (as well as loss and nonlinearity by the same argument) in the gain fibre, allowing us to simplify \eqref{eq:nlse} to
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A.
	\label{eq:gainexample}
\end{align}
To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, defined as
\begin{align}
	E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
and $\Es$ is the energy at which the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}, which depends on the power of the pump laser. Thus, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right) \right)^{1/2} A
	\label{eq:gainform}
\end{align}
for the amplification of the pulse within the Er-doped fibre.

Repeating this same procedure for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \exp( - \alpha L_T )A, \\
	D(A) &= \FTi{\exp( i \omega^2 \beta_2 L_D/2 ) \FT{A}}, \\
	F(A) &= A \exp( i \gamma |A|^2 L_f ),
\end{align}
where $R$ is the reflectivity of the optical coupler, $L_T$ is the total length of the laser circuit, $L_D$ is the `length' of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the optical coupler, and $\mathcal{F}$ denotes the Fourier transform. Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. This gives the freedom of choosing the profile of the modulation, how ever the operator sees fit. For simplicity, the form is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown that a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, it is common to use a saturable absorber in place of a modulator. In this case, we can exchange \eqref{eq:modform} for~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= \left( 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}} \right) A,
\end{align}
where $q_0$ is the modulation depth, $|A|^2$ is the instantaneous power of the pulse, and $P_\text{sat}$ is the saturation power.

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values, based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem, leaving only dimensionless equations with dimensionless parameters. The relative magnitudes of the dimensionless parameters characterize the dominant dynamics---like how fluid flow can be described by the Reynolds number. Each process in the laser can be better understood by re-scaling the time, energy, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

Returning to the process maps \eqref{eq:gainform}--\eqref{eq:modform}, we find the non-dimensional mappings to be (after dropping the tildes)
\begin{align}
	G(A) &= \left(E_g^{-1} W \left( a E_g \textrm{e}^{E_g}\right) \right)^{1/2} A, & F(A) &= \exp( i b |A|^2) A, \nonumber \\
	D(A) &= \FTi{\exp( i s^2 \omega^2 ) \FT{A}}, & L(A) &= h A, \label{eq:effects} \\
	M(A) &= \exp( -T^2 / 2 ) A, \nonumber
\end{align}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \exp( g_0 L_g ) \sim 8 \times 10^3, & 
		h &= (1 - R) \exp( - \alpha L_T) \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1, & s &= \left( \frac{\beta_2 L_D}{2 T_M^2} \right)^{1/2} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
as defined by the values in Table~\ref{tab:values}. Here, $a$ is a measure of the strength of the gain fibre, $h$ is the total loss due to absorption and the optical coupler, $b$ is the ratio of the frequency shift due to the Kerr nonlinearity to the width of the pulse, and $s$ is the ratio of the effectiveness of the CFBG to the width of the pulse. Furthermore, notice that in this non-dimensional form, the modulation has no associated parameter, and each of the other processes in \eqref{eq:effects} have their own independent, non-dimensional parameter.

% Table of parameter values
\begin{table*}[tbp]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline
		Parameter & Description & Value & Units \\
		\hline\noalign{\smallskip}
		$T_M$ & Characteristic modulation time & $15$--$150$ & ps \\
		$\beta_2 L_D$ & Dispersion provided by CFBG & $10$--$2000$ & ps$^2$ \\
		$\gamma$ & Kerr nonlinearity coefficient & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$L_f$ & Length of fibre between gain and optical coupler & $0.15$--$1$ & m \\
		$L_g$ & Length of the Er-doped gain fibre & $2$--$3$ & m \\
		$\alpha$ & Loss of fibre due to scattering and absorption & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy of Er-doped gain fibre & $10^3$--$10^4$ & pJ \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		$L_T$ & Total length of fibre in the cavity & $10$--$100$ & m \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

Finally, we briefly investigate the effect the order of the components has on the dynamics. In Section~\ref{sec:model}\ref{sec:effects} we justified the order of the gain, nonlinearity, and loss, but, just assumed dispersion preceded modulation. Here, we consider the effect of swapping the dispersion and modulation components. The energy landscape (Fig.~\ref{fig:energyswitch}) is largely the same as in Fig.~\ref{fig:energy}. In the previous case, with modulation following dispersion, more energy is removed from the pulse tails. Whereas, having modulation first, more of this energy remains in the cavity. This has the effect of making the laser more powerful than in the previous case, which in turn increases the size of the unstable region. Comparing with Fig.~\ref{fig:energy}, the unstable upper-left region spreads farther rightward, and the energy is larger in the stable region---especially away from the origin. Once again, the energy at the origin is 2.5487, as we are effectively removing the CFBG in the limit $s \rightarrow 0$.

@@

Portions of this work were presented at The V AMMCS International Conference in 2019, ``A new method of modelling tuneable lasers with functional composition'' \cite{metherallammcs}. BM acknowledges the support provided by the EPSRC Centre for Doctoral Training in Industrially Focused Mathematical Modelling (EP/L015803/1). The authors would like to thank C. Breward for critical reading of the manuscript.
